---
title: "Getting started with GroupCONSORT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with GroupCONSORT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GroupCONSORT)
library(dplyr)

# Clean up PNGs written by save_consort_plot() when the vignette finishes
withr::defer(
  file.remove(list.files(pattern = "\\.png$", full.names = FALSE)),
  envir = knitr::knit_global()
)

# Helper: save a consort_grob and return the path for include_graphics
save_fig <- function(p, name, dpi = 150) {
  save_consort_plot(p, name, formats = "png", dpi = dpi)
  paste0(name, ".png")
}
```

GroupCONSORT tracks patient attrition through inclusion/exclusion steps and
produces publication-quality CONSORT flowcharts, with optional per-group
(site / country) breakdowns. The API is a simple pipe chain — each
`include_if()` call filters the data and records what was dropped.

## Data

We use `survival::cgd`, a real placebo-controlled trial that ships with every
R installation. `prep_cgd_example()` reduces it to one row per patient and
adds a `region` column (North America / Europe).

```{r data}
cgd <- prep_cgd_example()
glimpse(cgd)
```

## Building a cohort

Start with `new_cohort()`, then pipe `include_if()` calls. The condition is
any ordinary dplyr filter expression.

```{r cohort}
cohort <- cgd |>
  new_cohort(label = "Randomised", id_col = "id", group_col = "region") |>
  include_if(age >= 5,           "Age >= 5 years") |>
  include_if(weight >= 15,       "Weight >= 15 kg") |>
  include_if(steroids == 0,      "Not on corticosteroids") |>
  include_if(follow_up_days > 0, "Positive follow-up time")

cohort
```

`get_tracker()` returns the underlying tibble — one row per group × step,
recording how many patients remain and how many were dropped.

```{r tracker}
get_tracker(cohort)
```

`get_data()` returns the filtered data frame ready for analysis.

```{r nrow}
nrow(get_data(cohort))
```

## Drawing the flowchart

`consort_plot()` returns a `consort_grob` object (a grid graphics tree).
Use `save_consort_plot()` to write it to disk at content-fitted dimensions,
then display it with `knitr::include_graphics()`. This pattern ensures the
figure is never stretched or compressed by knitr's default sizing.

```{r plot, fig.alt = "CONSORT flowchart showing attrition through four eligibility steps, with North America and Europe shown separately."}
p <- consort_plot(cohort)
knitr::include_graphics(save_fig(p, "elig"))
```

## Relabelling for publication

Rename steps and groups for display without touching the tracker. Only the
labels you want to change need to be listed.

```{r relabel, fig.alt = "CONSORT flowchart with relabelled steps and groups."}
p <- consort_plot(
  cohort,
  step_labels  = c("Not on corticosteroids" = "No concurrent corticosteroid use"),
  group_labels = c("North America" = "N. America")
)
knitr::include_graphics(save_fig(p, "label"))
```

## N/A cells for multi-source studies

When combining data from multiple sources, some steps may not apply to all
groups. Pass a data frame with `step` and `group` columns to `na_cells` —
those exclusion boxes show "N/A" while the main count boxes always show
totals.

```{r na-cells, fig.alt = "CONSORT flowchart with N/A in the exclusion box for North America at the weight step."}
cohort_multi <- cgd |>
  new_cohort("Randomised", id_col = "id", group_col = "region") |>
  include_if(age >= 5,                                  "Age >= 5 years") |>
  include_if(weight >= 15 | region == "North America",  "Weight >= 15 kg") |>
  include_if(steroids == 0,                             "Not on corticosteroids") |>
  include_if(follow_up_days > 0,                        "Positive follow-up time")

p <- consort_plot(
  cohort_multi,
  na_cells = data.frame(step = "Weight >= 15 kg", group = "North America")
)
knitr::include_graphics(save_fig(p, "na"))
```

## Merging trackers from separate datasets

When data from different sources are processed independently — for example,
countries run through separate pipelines where not all eligibility criteria
apply to every dataset — use `merge_trackers()` to combine them into a single
flowchart. Groups missing a step are carried forward at their last known count
and automatically flagged as N/A in the exclusion boxes.

`merge_trackers()` returns a list with `$tracker` and `$na_cells`, both of
which are passed directly to `consort_plot()`.

```{r merge, fig.alt = "CONSORT flowchart merging two independently-processed regional datasets, with N/A where a step did not apply."}
# Europe: both steps applied
tracker_eu <- cgd |>
  filter(region == "Europe") |>
  new_cohort("Randomised", id_col = "id", group_col = "region") |>
  include_if(age >= 5,     "Age >= 5 years") |>
  include_if(weight >= 15, "Weight >= 15 kg") |>
  get_tracker()

# North America: weight step was not run for this dataset
tracker_na <- cgd |>
  filter(region == "North America") |>
  new_cohort("Randomised", id_col = "id", group_col = "region") |>
  include_if(age >= 5, "Age >= 5 years") |>
  get_tracker()

result <- merge_trackers(tracker_eu, tracker_na)

p <- consort_plot(result$tracker, na_cells = result$na_cells)
knitr::include_graphics(save_fig(p, "merge"))
```

## Single-group studies

Set `group_col = NULL` for studies without a meaningful site grouping.

```{r single, fig.alt = "CONSORT flowchart with no per-group breakdown."}
p <- cgd |>
  new_cohort("Randomised", id_col = "id", group_col = NULL) |>
  include_if(age >= 5,           "Age >= 5 years") |>
  include_if(weight >= 15,       "Weight >= 15 kg") |>
  include_if(steroids == 0,      "Not on corticosteroids") |>
  include_if(follow_up_days > 0, "Positive follow-up time") |>
  consort_plot()
knitr::include_graphics(save_fig(p, "single"))
```

## Pre-filtering outside the pipeline

Complex criteria can be computed separately and joined back in before piping.

```{r prefilter}
xlinked_ids <- cgd |> filter(inherit == "X-linked") |> pull(id)

cgd |>
  new_cohort("Randomised", id_col = "id", group_col = "region") |>
  include_if(id %in% xlinked_ids, "X-linked inheritance") |>
  include_if(steroids == 0,       "Not on corticosteroids") |>
  get_tracker()
```

## Per-patient grouped filtering

When rows represent observations rather than patients (e.g. one row per ICU
admission), conditions involving aggregate functions like `min()` or `max()`
need to be evaluated per patient rather than across the whole dataset.
`include_if_grouped()` handles this by grouping on `id_col` before filtering,
so the aggregation is always patient-scoped.
```{r grouped-filter}
# prep_cgd_example() already has one row per patient, so we simulate a
# multi-row-per-patient dataset by duplicating with a fake admission time.
set.seed(42)
cgd_multi <- prep_cgd_example() |>
  dplyr::slice(rep(seq_len(dplyr::n()), each = 2)) |>
  dplyr::mutate(
    admission_day = follow_up_days + sample(c(0L, 5L, 10L), dplyr::n(),
                                            replace = TRUE)
  )

cohort_grouped <- cgd_multi |>
  new_cohort("All admissions", id_col = "id", group_col = "region") |>
  include_if(age >= 5, "Age >= 5 years") |>
  include_if_grouped(
    admission_day == min(admission_day),
    "First admission per patient"
  )

get_tracker(cohort_grouped)
```

If a patient has ties on the condition (two rows sharing the minimum value),
both rows are retained. Use `dplyr::slice_min()` directly on `get_data()` if
you need strict deduplication.

## Mutating between filter steps

`update_data()` replaces `cohort$data` without touching the tracker. This is
useful when you need to derive a column — via a data.table join, `mutate()`,
or any other operation — that cannot be expressed as a simple filter condition.
```{r update-data}
cohort_bmi <- cgd |>
  new_cohort("Randomised", id_col = "id", group_col = "region") |>
  include_if(age >= 5, "Age >= 5 years")

# Derive a column outside the pipe, then re-enter
enriched <- get_data(cohort_bmi) |>
  dplyr::mutate(bmi_proxy = weight / sqrt(age))

cohort_bmi <- update_data(cohort_bmi, enriched) |>
  include_if(bmi_proxy < 20, "BMI proxy < 20")

get_tracker(cohort_bmi)
```

## Saving

`save_consort_plot()` computes figure dimensions from the plot content so
text and boxes stay consistently sized regardless of how many steps or groups
there are. Figures are saved with a transparent background.

```{r save, eval = FALSE}
p <- consort_plot(cohort)
save_consort_plot(p, "consort_flowchart", formats = c("png", "pdf"))
```

The `scale` argument (default `1`) multiplies the natural mm dimensions —
increase it to produce a physically larger figure without changing layout
proportions. The `font_size` argument to `consort_plot()` scales all text
and box dimensions together.

## Layout stress tests

These chunks verify the layout holds up under edge cases — many groups,
many steps, long labels, and large font.

```{r stress-setup, include = FALSE}
set.seed(1)
df_many <- data.frame(
  id   = 1:120,
  site = rep(paste("Site", LETTERS[1:6]), each = 20),
  age  = sample(1:80, 120, replace = TRUE),
  wt   = sample(c(10, 20), 120, replace = TRUE)
)
```

```{r stress-many-groups, fig.alt = "Stress test: 6 groups, 2 steps."}
p <- df_many |>
  new_cohort("Screened", id_col = "id", group_col = "site") |>
  include_if(age >= 18, "Age >= 18 years") |>
  include_if(wt  >= 15, "Weight >= 15 kg") |>
  consort_plot()
knitr::include_graphics(save_fig(p, "six_groups"))
```

```{r stress-many-steps, fig.alt = "Stress test: 6 steps, 2 groups."}
p <- df_many |>
  dplyr::mutate(site = ifelse(site %in% c("Site A", "Site B", "Site C"),
                              "North", "South")) |>
  new_cohort("Screened", id_col = "id", group_col = "site") |>
  include_if(age >= 18,  "Age >= 18 years") |>
  include_if(age <= 75,  "Age <= 75 years") |>
  include_if(wt  >= 15,  "Weight >= 15 kg") |>
  include_if(wt  <= 100, "Weight <= 100 kg") |>
  include_if(age >= 20,  "Age >= 20 years") |>
  include_if(wt  >= 10,  "Weight >= 10 kg") |>
  consort_plot()
knitr::include_graphics(save_fig(p, "six_steps"))
```

```{r stress-long-labels, fig.alt = "Stress test: long step labels."}
p <- df_many |>
  dplyr::mutate(site = ifelse(site %in% c("Site A", "Site B", "Site C"),
                              "Northern Region", "Southern Region")) |>
  new_cohort("All screened participants at enrolment",
             id_col = "id", group_col = "site") |>
  include_if(age >= 18,
    "Participants aged 18 years or older at time of screening visit") |>
  include_if(wt >= 15,
    "Body weight at least 15 kg as recorded at baseline assessment") |>
  consort_plot()
knitr::include_graphics(save_fig(p, "long_label"))
```

```{r stress-large-font, fig.alt = "Stress test: font_size = 1.4."}
p <- df_many |>
  new_cohort("Screened", id_col = "id", group_col = "site") |>
  include_if(age >= 18, "Age >= 18 years") |>
  include_if(wt  >= 15, "Weight >= 15 kg") |>
  consort_plot(font_size = 1.4)
knitr::include_graphics(save_fig(p, "large_font"))
```
